---
title: "Exploration of Orca Data 2"
output: html_notebook
---

# Set Up

```{r Packages}
library(here)
library(tidyverse)
library(dplyr)
library(lme4)
library(stats)
#library(DMwR)
```

```{r Data Loading}
#Reading in the Data
library(here)
here()
orca_data<- read.csv(here("stredulinsky2021.csv")) #Kaylie/Clara/Madeline Read in 

```

```{r Data Exploration}
# Exploring
head(orca_data)
glimpse(orca_data)
#okay 25 columns that is a lot.

which(is.na(orca_data))
#no NA's
```

# Data Selection

Okay just looking at this there is a lot of data we don't need. I'm gonna make a subset of the data with the columns we are looking for. 

Pod: Whole group, we are just looking at pod


## Column selection

### Initial Hypothesis (food and relatedness with pod cohesion)

- Year
- Group ID
- CK.oa (Chinook salmon ocean abundance index)
- CK.tr (Chinook salmon terminal run reconstruction abundance index) (accounting for industrial fishers affecting abundance)
- CM (chum salmon terminal run abundance index)
- DPERg (Total estimated daily prey energetic requirement of the pod)
- FLg (number of lactating females in the pod (proportion to pod size))
- FRg (number of reproductive age females in pod)
- Ng (number of individuals in pod)

This will give me 8 columns instead of 25

#### Subsetting the Data

```{r Selecting Columns 1}
#Creating Subset
orca_data_subset <- select(orca_data, Year, Group.ID,
                           CK.oa, CK.tr, CM, DPERg, FLg, FRg, Ng)
                          

head(orca_data_subset)
```

```{r NA Check}
#Checking for Na's
which(is.na(orca_data_subset))
#okay so it appears we don't have any missing values. 
```

### Secondary Hypothesis - Food and Fecundity

```{r Selecting Columns 2}
orca_data_sub2 <- orca_data %>%
  select(Year, Group.ID, CK.oa, CK.tr, CM, DPERg, FLg, Ng)

#selecting 6 random groups to look at because 56 is hard to see at once
groups <- sample(1:56, 6, replace = F) #sample 6 random numbers from 1-56
groups <- as.character(groups)
for(i in 1:length(groups)){ #for each number
  if(nchar(groups[i])==1){ #if the number of characters is 1 (1, 2, 3...)
    groups[i] <- paste("G00", groups[i], sep="") #add G00 in front
  }
  else{ #if the number is not 1 (10, 11, 12...)
  groups[i] <- paste("G0", groups[i], sep="") #add G0 in front
  }
}

orca_data_sub2_6 <- orca_data_sub2 %>% #filter for the 6 groups
  filter(Group.ID == groups[1] | Group.ID == groups[2] | Group.ID == groups[3] |
         Group.ID == groups[4] | Group.ID == groups[5] | Group.ID == groups[6])

#Checking number of observations from each group
orca_data_sub2 %>% 
  group_by(Group.ID) %>% 
  tally() %>% 
  arrange(n, desc(n))

#G002, G003 have 1 observation
#G005, G0019, G022, G024, G027, G037, G046 have 2 observations
```

# We need to Remove groups with less then ?


# Plotting


## Secondary Hypothesis

```{r Exploratory Graphing 2}
#Salmon over time
ggplot(orca_data_sub2, aes(Year, CK.oa))+
  geom_point()+
  geom_line()

ggplot(orca_data_sub2, aes(Year, CK.tr))+
  geom_point()+
  geom_line()

ggplot(orca_data_sub2, aes(Year, CM))+
  geom_point()+
  geom_line()

#Lactating females over time in groups
ggplot(orca_data_sub2, aes(Year, FLg))+
  geom_point()+
  facet_wrap(~Group.ID)

ggplot(orca_data_sub2_6, aes(Year, FLg))+
  geom_point()+
  geom_line()+ #improves readability
  facet_wrap(~Group.ID)


#Number of individuals in groups
ggplot(orca_data_sub2, aes(Year, Ng))+
  geom_point()+
  facet_wrap(~Group.ID)

ggplot(orca_data_sub2_6, aes(Year, Ng))+
  geom_point()+
  geom_line()+
  facet_wrap(~Group.ID)

#pod energetic requirement
ggplot(orca_data_sub2, aes(Year, DPERg))+
  geom_point()+
  facet_wrap(~Group.ID)

ggplot(orca_data_sub2_6, aes(Year, DPERg))+
  geom_point()+
  geom_line()+
  facet_wrap(~Group.ID)
```

## Testing Normality

```{r}
ggplot(orca_data_sub2, aes(FLg))+
  geom_histogram() #histogram of FLg to assess normality
```
## Unscaling

```{r}
# value = x-mean(x) / 2sd(x)
# descale <- function(x, sd){
#   vals <- c()
#   for(i in 1:length(x)){
#     #vals[i] <- (x[i]-mean(x))/(2*sd)
#     vals[i] <- (x[i]+mean(x))*(2*sd)
#   }
#   return(vals)
# }
# 
# 
# unscaledFLg <- descale(orca_data_sub2$FLg, 0.73)
# 
# unscaledData <- orca_data_sub2 %>% 
#   select(FLg) %>% 
#   unscale()
# 
# unscaledFLg <- unscale(2*orca_data_sub2$FLg)
# unscaledCK.oa <- unscale(2*orca_data_sub2$CK.oa)
```


## Testing Assumptions (LMM)
1. Random Effects are Normally Distributed
2. Response (lactating Females) is normally distributed
3. Homogenity/Homoscedasticity of variance


Testing 1. Random Effects Normally distributed
```{r}
orca_data_sub2 %>% 
  select(CK.oa, Year, Group.ID) %>% 
  plot()
```

```{r}
orca_data_sub2 %>% 
  select(CK.oa) %>% 
  summary() 

hist(orca_data_sub2$CK.oa)
hist(orca_data_sub2$FLg) #Y is not normally distributed Transform to log?
```



###Making our Linear Model
Model #1 Treat Salmon Data as a fixed Variable. For the sake of just exploration I'm going to use just CK.oa for now

Random Variables; Year, Group ID, 
Fixed Variables : FLg, FRg, Salmon Data

#LMM with Randomized Slope
- we have random effects
- data is dependen on one another
- our data is not normal
- we expect the number of salmon to increase 'randomly' with number of lactating females

#CK.oa and FLg- simplest model just year as random effect
```{r}
library(lme4)
library(lmerTest)
CK_FLg_Year<-lmer(FLg~CK.oa+ (1|Year), data = orca_data_sub2, REML= FALSE)
summary (CK_FLg_Year)
```
__BIGNOTE___ (we dont actually know what salmon data units is so this means nothing rn :)


#CK.Oa, FLG ; Random effects: year and Group ID
```{r}
CK_FLg_Year_GroupID<-lmer(FLg~CK.oa + (1|Year) + (1|Group.ID), data = orca_data_sub2, REML= FALSE)
summary(CK_FLg_Year_GroupID)

plot(CK_FLg_Year_GroupID) # I dont know
```

I am going to do some transformations to see if I can make the plot less uggo
```{r}
# log transforming FLG 
orca_data_sub2$FLg <- log(orca_data_sub2$FLg)

CK_FLg_Year_GroupID_logtransform<-lmer(FLg~CK.oa + (1|Year) + (1|Group.ID), data = orca_data_sub2, REML= FALSE)
#looks better! 
plot(CK_FLg_Year_GroupID_logtransform)

#exponentially transforming FLG
orca_data_sub2$FLg <- exp(orca_data_sub2$FLg)


CK_FLg_Year_GroupID_exptransform<-lmer(FLg~CK.oa + (1|Year) + (1|Group.ID), data = orca_data_sub2, REML= FALSE)
#this also looks somewhat better than the original but I think the log one is a better transformation 
plot(CK_FLg_Year_GroupID_exptransform)
```


Kaylie's Attempt at Generating Data

Okay so based on Salmon data collection from 
https://www.npafc.org/wp-content/uploads/Public-Documents/2009/1199Rev1WGSA.pdf
pg - 51 (BC Chum Salmon)

For each year from 1980 to 2010 need to randomly select a salmon quantity.
Hypothetically, the abundance would go down towards the 2000's but for this beginning function I'm just gonna have it choose randomly.

Data set above:
Min value: 0.46 (million)
Max Value: 6.17 (million)

## Chum Generator
```{r}

years_measured<-(c(seq(1980,2010,1)))
#Numbers are in values of Millions of Fish
chum_generator <- function(){
  chum_density <- sample(x = seq(0.46,6.17,0.02), size = length(years_measured), replace = T) #choosing a random value between the range of 0.46 to 6.17, Replacing is True
  yearly_chum <- data.frame(years_measured, chum_density) #making data frame for each year and with the random density
    return(yearly_chum)
  }
  
chum_generator()

#Okay for reference, sample needs the x (the range to pick from) and then the size (which is how many variables its gonna pick); so don't need a for loop
  
```

## Chinook Generator
Based on https://www.npafc.org/wp-content/uploads/Public-Documents/2009/1199Rev1WGSA.pdf
pg 123
From 1980- 2008 (in millions of fish)
Min value: 0.13
Max Value: 1.27

```{r}

#Numbers are in values of Millions of Fish
chinook_generator <- function(){
  chinook_density <- sample(x = seq(0.13,1.27,0.02), size = length(years_measured), replace = T) #choosing a random value between the range of 0.46 to 6.17, Replacing is True
  yearly_chinook <- data.frame(years_measured, chinook_density) #making data frame for each year and with the random density
    return(yearly_chinook)
  }
  
chinook_generator()

```

#Generator for Salmon in DF format
```{r}
salmon <- data.frame(chinook_generator(), chum_generator())

salmon_data<-salmon %>% 
  select(years_measured,chinook_density,chum_density) %>% 
  mutate(total_salmon = chinook_density + chum_density) 
```

```{r}
hist(salmon_data$total_salmon)
```


#Pod Size Generator 
Based on 

```{r}
group_number<-seq(1,16,1)
pod <- function(){
  n<- rpois(1, lambda = 300)
  pod_size<-rmultinom(1, size = n, prob = rep(1,16)) 
  #multinorm to make sure orcas add to 300, repeat 16 times
  pod_size_by_group <- data.frame(group_number,pod_size ) #puttin gin dataframe
    return(pod_size_by_group)
  }
 
pod()

```

## Amount of Reproductive females
Based on (INSERt link)
```{r}
pod_data<- pod() %>% 
  mutate(number_repro_females = round(pod_size * 0.22,0))
#rounding upwards to get whole number

```

# Number of Lactating Females
```{r}
Lactating_females <- c()
for(i in 1:nrow(pod_data)){
  Lactating_females[i] <- rbinom(1,pod_data$number_repro_females[i], prob = 0.2)
}

# Lactating_females

pod_data_2<-pod_data %>% 
  mutate(Lactating_females = Lactating_females)

# pod_data_2
```

```{r}
#n <- 31
data_Gen <- function(beta1, beta2){
  pod_data_Year <- data.frame()
  
  #Salmon Data
  #generate salmon data in a dataframe
  salmon <- data.frame(chinook_generator(), chum_generator())
  
  salmon_data<-salmon %>%  #format data frame, add total salmon
    select(years_measured,chinook_density,chum_density) %>% 
    mutate(total_salmon = chinook_density + chum_density) 
  
  #fitting the salmon data to the orca data
  rep_salmon_data <- data.frame(chinook_density = c(), 
                                chum_density = c(), 
                                total_salmon = c())
  
  x <- 1
  i <- 1
  while(x < 496){
    #for each 16 rows of the empty data frame, add the corresponding year 
    #from salmon data
    rep_salmon_data[c(x:(x+16)),1] <- salmon_data[i,2]
    rep_salmon_data[c(x:(x+16)),2] <- salmon_data[i,3]
    rep_salmon_data[c(x:(x+16)),3] <- salmon_data[i,4]
    x <- x+16 #increase rows in repeated salmon data
    i <- i+1 #increase row in salmon data
  }
  rep_salmon_data <- rep_salmon_data[-c(497),]
  
  colnames(rep_salmon_data) <- c("chinook_abundace", "chum_abundance", "total_salmon")
  
  # Orca Data Generation
  group_number<-seq(1,16,1)
  n <- 31
  
  for(i in 1:n){ #generate pod sizes
    pod <- function(){
      #poisson distribution to choose total population each year
      s <- rpois(1, lambda = 300)
      #multinomial distribution keeps orcas at total pop size, make 16 pods
      pod_size <- rmultinom(1, size = s, prob = rep(1,16))
      year <- i+1979
      pod_size_by_group <- data.frame(year, group_number, pod_size)
      return(pod_size_by_group)
    }

    # make the number of lactating females
    pod_data <- pod() #call pod function above to make pods
    # %>% mutate(number_repro_females = round(pod_size * 0.22,0)) - *removed*
    Lactating_females <- c()
    for(i in 1:nrow(pod_data)){ #calculate lactating females
      #beta - makes lactating females dependent on salmon + pod size
      Lactating_females[i] <- rpois(1,
                                lambda = beta1*(pod_data$pod_size[i])+ 
                                         beta2*rep_salmon_data$total_salmon[i])
      #so: determine lactating females from pod size + additional amount 
      #w/ more salmon
      #
      # originally was: rbinom(1, pod_data$number_repro_females, prob = 0.2)
    }
    pod_data_2<-pod_data %>% 
      mutate(lactating_females = Lactating_females)
    pod_data_Year <- rbind(pod_data_Year, pod_data_2)
  }

  #combining the dataframe
  gen_Data <- pod_data_Year %>%
    mutate(chinook_abundance = c(rep_salmon_data[,1])) %>%
    mutate(chum_abundance = c(rep_salmon_data[,2])) %>%
    mutate(total_salmon = c(rep_salmon_data[,3]))
  
  return(gen_Data)
}
#year_pod_data <- lactating_Orca_Data_Gen(31)

#fitting the generated salmon data to proper rows for orcas
# rep_salmon_data <- data.frame(chinook_density = c(), chum_density = c(), total_salmon = c())
# x <- 1
# i <- 1
# while(x < nrow(year_pod_data)){
#   #rep_salmon_data[c(x:(x+16)),1] <- salmon_data[i,1]
#   rep_salmon_data[c(x:(x+16)),1] <- salmon_data[i,2]
#   rep_salmon_data[c(x:(x+16)),2] <- salmon_data[i,3]
#   rep_salmon_data[c(x:(x+16)),3] <- salmon_data[i,4]
#   x <- x+16
#   i <- i+1
# }
# rep_salmon_data <- rep_salmon_data[-c(497),]
# 
# #combining the dataframe
# gen_Data <- year_pod_data %>%
#   mutate(chinook_abundance = c(rep_salmon_data[,1])) %>%
#   mutate(chum_abundance = c(rep_salmon_data[,2])) %>%
#   mutate(total_salmon = c(rep_salmon_data[,3]))
# 
# ggplot(gen_Data, aes(lactating_females)) +
#   geom_histogram()
# 
# ggplot(gen_Data, aes(lactating_females)) +
#   geom_histogram() +
#   facet_wrap(group_number)
```

MISC>
i think if salmon is a random variable it should be (1+ck.oa|Year) (because ck is dependent on y ear)

# Power Analysis

```{r, warning = F}
# beta1 <- 0.22 # seq(0.2, 0.24, 0.01)
# beta2 <- 0.1 # seq(-0.5, 0.5, 0.1)
# beta <- c(beta1, beta2)

params <- expand.grid(beta1 = seq(0.2,0.24,0.01), 
                      beta2 = c(0.0000001,0.00001,0.0001,0.001,0.01,0.1))
#beta2 = seq(0.01, 1, 0.01)) # - another option maybe?

#beta needs to be a vector with two entries, 1st beta1, 2nd beta2
power_poisson <- function(beta){
  
  sims <- 1000 #simulation number
  pValSalmon <- c()
  pValPod <- c()
  SalmonEffect <- c()
  PodEffect <- c()
  models <- NULL
  predictions <- NULL
  Intercept <- c()
  
  for(i in 1:sims){ #loop for the number of simulations
    #for(i in 1:nrow(beta)){
    data <- data_Gen(beta[1], beta[2]) #generate the orca data for 31 years (1980-2010)
    #data <- data_Gen(params[i,1], params[i,2])
    #make a glm with lactating females as (y), and fixed effects total_salmon and pod_size as (x). Random effects arr year and group_number
    model <- glmer(lactating_females~total_salmon+pod_size+
                   (1|group_number)+(1|year), 
                 family = poisson, data)
    #save the effect salmon has
    models[[i]] <- model
    #prediction[i] <- predict(model)
    
    predictions[[i]] <- cbind(lact_fem_predict=
                                predict(models[[i]], 
                                        newdata = data.frame(
                                          total_salmon=data$total_salmon,
                                          pod_size = data$pod_size,
                                          year=data$year,
                                          group_number=data$group_number)), 
                              total_salmon = data$total_salmon,
                              year = data$year,
                              pod_size = data$pod_size,
                              group_number = data$group_number,
                              index = i)
    
    SalmonEffect[i] <- summary(model)$coefficient[2,1]
    PodEffect[i] <- summary(model)$coefficient[3,1]
    
    Intercept[i] <- summary(model)$coefficient[1,1] #save intercept
    
    pValSalmon[i] <- summary(model)$coefficient[2,4]
    pValPod[i] <- summary(model)$coefficient[3,4] #save model significance
    #}
  }
  
  powEstSalmon <- length(which(pValSalmon<0.05))/length(pValSalmon)
  powEstPod <- length(which(pValPod<0.05))/length(pValPod)
  
  return(list(SalmonEffect, PodEffect, Intercept,
              powEstSalmon, powEstPod, do.call(rbind, predictions)))
}

#results <- power_poisson(beta)
#results <- power_poisson(params)
PowersSalmon <- c()
powEstPod <- c()
overall_Results <- NULL

for (i in 1:nrow(params)){
  beta <- c(params[i,1], params[i,2])
  result <- power_poisson(beta)
  overall_Results[i] <- result
  PowersSalmon[i] <- result[[4]]
  powEstPod[i] <- result[[5]]
}

print(paste("Salmon Power Estimate: ", overall_Results[4]))
minSaEffect <- min(as.data.frame(overall_Results[1]))
maxSaEffect <- max(as.data.frame(overall_Results[1]))
print(paste("Range of Salmon Effects: ", minSaEffect, "-", maxSaEffect))
print(paste("Pod Power Estimate: ", overall_Results[5]))
minPodEffect <- min(as.data.frame(overall_Results[2]))
maxPodEffect <- max(as.data.frame(overall_Results[2]))
print(paste("Range of Pod Size Effects: ", minPodEffect, "-", maxPodEffect))

# cbind(SalmonEffect = results[[1]]) %>%
#   ggplot(aes(x = SalmonEffect))+
#   geom_histogram()

# PowersSalmon <- c()
# powEstPod <- c()
# 
# for (i in 1:nrow(params)){
#   beta <- c(params[i,1], params[i,2])
#   result <- power_poisson(beta)
#   PowersSalmon[i] <- result[[4]]
#   powEstPod[i] <- result[[5]]
# }

cbind(PowersSalmon, powEstPod, params) # Creating Dataframe of Powers at various Beta levels. 

seq(min(as.data.frame(result[[6]])$total_salmon),
    max(as.data.frame(result[[6]])$total_salmon), length = 1000) -> salmons

seq(min(as.data.frame(result[[6]])$pod_size),
    max(as.data.frame(result[[6]])$pod_size), length = 1000) -> pods

salmon_lines <- NULL
pod_lines <- NULL

for (i in 1:length(result[[3]])){
  lines1 <- as.data.frame(cbind(salmons = salmons)) %>%
    mutate(lact_prediction = result[[3]][i]+
             result[[1]][i]*salmons)
  lines1 <- cbind(lines1, index_model = i)
  salmon_lines <- rbind(salmon_lines, lines1)
  # POD SIZE IS CONSTANT HERE
  
  lines2 <- as.data.frame(cbind(pods = pods)) %>%
    mutate(lact_prediction = result[[3]][i]+
             result[[2]][i]*pods)
  lines2 <- cbind(lines2, index_model = i)
  pod_lines <- rbind(pod_lines, lines2)
  # SALMON CONSTANT
}




as.data.frame(salmon_lines) %>%
  ggplot(aes(x = salmons, y = lact_prediction, group = index_model))+
  geom_line(color = "gray")+
  geom_line(aes(x = mean(salmons), y = mean(lact_prediction)), colour = "blue")+
  labs(x = "Salmon Abundance (millions of fish)", y ="Predicted Number of Lactating Females per Pod" , title = "Effect of Salmon Abundance on Lactating Females")+
  theme_bw()


#Making Average line - I need an average intercept
mean(pod_lines$pods) #22
mean(pod_lines$lact_prediction) #1.717725
#Okay so y = mx+b

as.data.frame(pod_lines) %>% 
  ggplot(aes(x = pods, y = lact_prediction, group = index_model))+
  geom_line(color = "gray")+
  labs(x = "Size of Pod", y = "Predicted Number of Lactating Females", title = "Pod Size on Number of Lactating Females")+
  theme_bw()
```

```{r}
library(tidyverse)
library(here)

orcaCensus <- read.csv(here("NRKW Population 1.csv"))
#orcapop<-read.csv("C:/Users/Owner/Desktop/fall 23'/EEB 313/NRKW Population 1.csv")

library(ggplot2)
 
ggplot(data = orcapop, aes(x = Females_fecund_available)) +
  geom_histogram()

ggplot (data = orcapop, aes(x = Pregnancy_count)) +
  geom_histogram()

#measure of lactating 

ggplot(data = orcapop, aes(x = Females_fecund_unavailable)) +
  geom_histogram()

ggplot(data = orcapop, aes(x = Birth_count)) +
  geom_histogram()


```
#GLMM / LMM

```{r}
# (val - mean) / 2*SD
# mean_Size <- mean(orcaCensus$Size_best)
# sd_Size <- sd(orcaCensus$Size_best)
# transformed_Total <- (orcaCensus$Size_best-mean_Size)/(2*sd_Size)
# untransformed_Group_Num <- NULL
# for(i in 1:564){
#   untransformed_Group_Num[i] <- (orca_data$Ng[i]*(2*sd_Size))+mean_Size
# }
# mean_Fecund_Unav <- mean(orcaCensus$Females_fecund_unavailable)
# sd_Fecund_Unav <- sd(orcaCensus$Females_fecund_unavailable)
# untransformed_FLg <- NULL
# for(i in 1:564){
#   untransformed_FLg[i] <- (orca_data$FLg[i]*(2*sd_Fecund_Unav))+mean_Fecund_Unav
# }
#untransformed_Group_Num <- (orca_data$Ng/sD_Size)+mean_Size
# THIS GIRL SHOUD NEVER WRITE A PAPER AGAIN

pods_Year <- data.frame()
for(i in 1:31){
  year <- i+1979
  pods_Year[i,1] <- year
  pods_Year[i,2] <- orca_data %>% tally(Year == year)
  pods_Year[i,3] <- (orcaCensus$Size_best[orcaCensus$YEAR==(year)]/pods_Year[i,2])
}
colnames(pods_Year) <- c("Year", "Pod_Count", "Avg_Orcas_Per_Pod")

```

From Fig. 1c:
1980: 20 pods
1981: 18 pods
1982: 18 pods
1983: 20 pods
1984: 21 pods
1985: 17 pods
1986
1990: 21 pods
1995: 25 pods
2000: 24 pods
2005: 23 pods
2010: 24 pods



