---
title: "Exploration of Orca Data"
output: html_notebook
---

# Set Up

```{r Packages}
library (here)
library(tidyverse)
library(dplyr)
```

```{r Data Loading}
#Reading in the Data
here()
orca_data <- read.csv(here("Project", "Raw Data", "stredulinsky2021.csv"))
```

```{r Data Exploration}
# Exploring
head(orca_data)
glimpse(orca_data)
#okay 25 columns that is a lot.

which(is.na(orca_data))
#no NA's
```

# Data Selection

Okay just looking at this there is a lot of data we don't need. I'm gonna make a subset of the data with the columns we are looking for. (I'm not sure if it would be better to keep them all and do a mixed effects yet.)

Pod: Whole group
Subunit: within pod (different females with kids therfore the subunit is the largest familial group)

## Column selection

### Initial Hypothesis (food and relatedness with pod cohesion)

- Year
- Group ID
- Submatriarch ID
- HWI (Half weight index value estimating the pod’s cohesion in the given year)
- CK.oa (Chinook salmon ocean abundance index)
- CK.tr (Chinook salmon terminal run reconstruction abundance index) (accounting for industrial fishers affecting abundance)
- CM (chum salmon terminal run abundance index)
- DPERg (Total estimated daily prey energetic requirement of the pod)
- DPERs (Total estimated daily prey energetic requirement of the subunit)
- Lm (Indicates whether a maternal ancestor common to all pod members is alive (1) or dead (0) in given year
- Ng (number of individuals in pod)
- Ns (number of individuals in subunit)
- Rg (Average pairwise maternal relatedness in the pod, corrected for pod size
- Rsm Average pairwise maternal relatedness between the submatriarch and her
descendants (the subunit), relative to that between her and the rest of her pod
- Sg (Number of complete units in the pod)

This will give me 14 columns instead of 25

#### Subsetting the Data

```{r Selecting Columns}
#Creating Subset
orca_data_subset <- select(orca_data, Year, Group.ID, Submatriarch.ID,
                           HWI, CK.oa, CK.tr, CM, DPERg, DPERs, Lm, Ng, 
                           Ns, Rg, Rsm, Sg)

head(orca_data_subset)
```

```{r NA Check}
#Checking for Na's
which(is.na(orca_data_subset))
#okay so it appears we don't have any missing values. 
```

### Secondary Hypothesis - Food and Fecundity

```{r Selecting Columns 2}
orca_data_sub2 <- orca_data %>%
  select(Year, Group.ID, Ag, PMg, CK.oa, CK.tr, DPERg, FLg)

#selecting 6 random groups to look at because 56 is hard to see at once
groups <- sample(1:56, 6) #sample 6 random numbers from 1-56
groups <- as.character(groups)
for(i in 1:length(groups)){ #for each number
  if(nchar(groups[i])==1){ #if the number of characters is 1 (1, 2, 3...)
    groups[i] <- paste("G00", groups[i], sep="") #add G00 in front
  }
  else{ #if the number is not 1 (10, 11, 12...)
  groups[i] <- paste("G0", groups[i], sep="") #add G0 in front
  }
}
orca_data_sub2_5 <- orca_data_sub2 %>% #filter for the 6 groups
  filter(Group.ID == groups[1] | Group.ID == groups[2] | Group.ID == groups[3] |
         Group.ID == groups[4] | Group.ID == groups[5] | Group.ID == groups[6])
```

# Plotting

## Initial Hypothesis

```{r Exploratory Plotting}
#Year on cohesion
ggplot(data = orca_data_subset, aes(x = Year, y = HWI))+
  geom_point()+
  facet_wrap(~Group.ID)
# Chinook abundance on cohesion
ggplot(data = orca_data_subset, aes(x = CK.tr, y = HWI))+
  geom_point()+
  facet_wrap(~Group.ID)
# Maternity (by pod) on cohesion
ggplot(data = orca_data_subset, aes (x = Rg, y = HWI))+
  geom_point()+
  facet_wrap(~Group.ID)


```

## Secondary Hypothesis

```{r Exploratory Graphing 2}
#Salmon over time
ggplot(orca_data_sub2, aes(Year, CK.oa))+
  geom_point()

#Lactating females over time in groups
ggplot(orca_data_sub2, aes(Year, FLg))+
  geom_point()+
  facet_wrap(~Group.ID)

ggplot(orca_data_sub2_5, aes(Year, FLg))+
  geom_point()+
  geom_line()+ #improves readability
  facet_wrap(~Group.ID)

#Median age in groups
ggplot(orca_data_sub2, aes(Year, Ag))+
  geom_point()+
  facet_wrap(~Group.ID)

ggplot(orca_data_sub2_5, aes(Year, Ag))+
  geom_point()+
  geom_line()+ #improves readability
  facet_wrap(~Group.ID)
```


# Analysis

ANALYSIS PLAN

I want to view the data so I'm going to make a linear mixed model.
- Random Variables:
  - Year
  - CK.oa
  - CK.tr
  - CM
   - Note turn Group.ID and Submatriarch into as.factor

- Fixed variables: 
  - HWI
  - DPERg
  - DPERs

Not Effects?
Group.ID
Submatriarch.ID

```{r}

```


Things we need to decide
- pod or subunit
- chinook salmon vs chum or both?
- Rsm vs Rg











