---
title: "Exploration of Orca Data 2"
output: html_notebook
---

# Set Up

```{r Packages}
library(here)
library(tidyverse)
library(dplyr)
```

```{r Data Loading}
#Reading in the Data
library(here)
here()
#orca_data <- read.csv(here("Project", "Raw Data", "stredulinsky2021.csv")) 
# Maddy's read in
orca_data<- read.csv(here("stredulinsky2021.csv")) #Kaylie's'/Clara Read in 

```

```{r Data Exploration}
# Exploring
head(orca_data)
glimpse(orca_data)
#okay 25 columns that is a lot.

which(is.na(orca_data))
#no NA's
```

# Data Selection

Okay just looking at this there is a lot of data we don't need. I'm gonna make a subset of the data with the columns we are looking for. 

Pod: Whole group, we are just looking at pod


## Column selection

### Initial Hypothesis (food and relatedness with pod cohesion)

- Year
- Group ID
- CK.oa (Chinook salmon ocean abundance index)
- CK.tr (Chinook salmon terminal run reconstruction abundance index) (accounting for industrial fishers affecting abundance)
- CM (chum salmon terminal run abundance index)
- DPERg (Total estimated daily prey energetic requirement of the pod)
- FLg (number of lactating females in the pod (proportion to pod size))
- FRg (number of reproductive age females in pod)
- Ng (number of individuals in pod)

This will give me 8 columns instead of 25

#### Subsetting the Data

```{r Selecting Columns 1}
#Creating Subset
orca_data_subset <- select(orca_data, Year, Group.ID,
                           CK.oa, CK.tr, CM, DPERg, FLg, FRg, Ng)
                          

head(orca_data_subset)
```

```{r NA Check}
#Checking for Na's
which(is.na(orca_data_subset))
#okay so it appears we don't have any missing values. 
```

### Secondary Hypothesis - Food and Fecundity

```{r Selecting Columns 2}
orca_data_sub2 <- orca_data %>%
  select(Year, Group.ID, CK.oa, CK.tr, CM, DPERg, FLg, Ng)

#selecting 6 random groups to look at because 56 is hard to see at once
groups <- sample(1:56, 6, replace = F) #sample 6 random numbers from 1-56
groups <- as.character(groups)
for(i in 1:length(groups)){ #for each number
  if(nchar(groups[i])==1){ #if the number of characters is 1 (1, 2, 3...)
    groups[i] <- paste("G00", groups[i], sep="") #add G00 in front
  }
  else{ #if the number is not 1 (10, 11, 12...)
  groups[i] <- paste("G0", groups[i], sep="") #add G0 in front
  }
}

orca_data_sub2_6 <- orca_data_sub2 %>% #filter for the 6 groups
  filter(Group.ID == groups[1] | Group.ID == groups[2] | Group.ID == groups[3] |
         Group.ID == groups[4] | Group.ID == groups[5] | Group.ID == groups[6])

#Checking number of observations from each group
orca_data_sub2 %>% 
  group_by(Group.ID) %>% 
  tally() %>% 
  arrange(n, desc(n))

#G002, G003 have 1 observation
#G005, G0019, G022, G024, G027, G037, G046 have 2 observations
```

# We need to Remove groups with less then ?


# Plotting


## Secondary Hypothesis

```{r Exploratory Graphing 2}
#Salmon over time
ggplot(orca_data_sub2, aes(Year, CK.oa))+
  geom_point()+
  geom_line()

ggplot(orca_data_sub2, aes(Year, CK.tr))+
  geom_point()+
  geom_line()

ggplot(orca_data_sub2, aes(Year, CM))+
  geom_point()+
  geom_line()

#Lactating females over time in groups
ggplot(orca_data_sub2, aes(Year, FLg))+
  geom_point()+
  facet_wrap(~Group.ID)

ggplot(orca_data_sub2_6, aes(Year, FLg))+
  geom_point()+
  geom_line()+ #improves readability
  facet_wrap(~Group.ID)


#Number of individuals in groups
ggplot(orca_data_sub2, aes(Year, Ng))+
  geom_point()+
  facet_wrap(~Group.ID)

ggplot(orca_data_sub2_6, aes(Year, Ng))+
  geom_point()+
  geom_line()+
  facet_wrap(~Group.ID)

#pod energetic requirement
ggplot(orca_data_sub2, aes(Year, DPERg))+
  geom_point()+
  facet_wrap(~Group.ID)

ggplot(orca_data_sub2_6, aes(Year, DPERg))+
  geom_point()+
  geom_line()+
  facet_wrap(~Group.ID)
```

## Testing Assumptions (LMM)
1. Random Effects are Normally Distributed
2. Response (lactating Females) is normally distributed
3. Homogenity/Homoscedasticity of variance


Testing 1. Random Effects Normally distributed
```{r}
orca_data_sub2 %>% 
  select(CK.oa, Year, Group.ID) %>% 
  plot()
```

```{r}
orca_data_sub2 %>% 
  select(CK.oa) %>% 
  summary() 

hist(orca_data_sub2$CK.oa)
hist(orca_data_sub2$FLg) #Y is not normally distributed Transform to log?
```



###Making our Linear Model
Model #1 Treat Salmon Data as a fixed Variable. For the sake of just exploration I'm going to use just CK.oa for now

Random Variables; Year, Group ID, 
Fixed Variables : FLg, FRg, Salmon Data

#LMM with Randomized Slope
- we have random effects
- data is dependen on one another
- our data is not normal
- we expect the number of salmon to increase 'randomly' with number of lactating females

#CK.oa and FLg- simplest model just year as random effect
```{r}
library(lme4)
library(lmerTest)
CK_FLg_Year<-lmer(FLg~CK.oa+ (1|Year), data = orca_data_sub2, REML= FALSE)
summary (CK_FLg)
```
__BIGNOTE___ (we dont actually know what salmon data units is so this means nothing rn :)


#CK.Oa, FLG ; Random effects: year and Group ID
```{r}
CK_FLg_Year_GroupID<-lmer(FLg~CK.oa + (1|Year) + (1|Group.ID), data = orca_data_sub2, REML= FALSE)
summary (CK_FLg)

plot(CK_FLg_Year_GroupID) # I dont know
```
MISC>
i think if salmon is a random variable it should be (1+ck.oa|Year) (because ck is dependent on y ear)




