---
title: "Exploration of Orca Data 2"
output: html_notebook
---

# Set Up

```{r Packages}
library(here)
library(tidyverse)
library(dplyr)
library(ggfortify)
library(stats)
#library(DMwR)
```

```{r Data Loading}
#Reading in the Data
library(here)
here()
orca_data<- read.csv(here("stredulinsky2021.csv")) #Kaylie/Clara/Madeline Read in 

```

```{r Data Exploration}
# Exploring
head(orca_data)
glimpse(orca_data)
#okay 25 columns that is a lot.

which(is.na(orca_data))
#no NA's
```

# Data Selection

Okay just looking at this there is a lot of data we don't need. I'm gonna make a subset of the data with the columns we are looking for. 

Pod: Whole group, we are just looking at pod


## Column selection

### Initial Hypothesis (food and relatedness with pod cohesion)

- Year
- Group ID
- CK.oa (Chinook salmon ocean abundance index)
- CK.tr (Chinook salmon terminal run reconstruction abundance index) (accounting for industrial fishers affecting abundance)
- CM (chum salmon terminal run abundance index)
- DPERg (Total estimated daily prey energetic requirement of the pod)
- FLg (number of lactating females in the pod (proportion to pod size))
- FRg (number of reproductive age females in pod)
- Ng (number of individuals in pod)

This will give me 8 columns instead of 25

#### Subsetting the Data

```{r Selecting Columns 1}
#Creating Subset
orca_data_subset <- select(orca_data, Year, Group.ID,
                           CK.oa, CK.tr, CM, DPERg, FLg, FRg, Ng)
                          

head(orca_data_subset)
```

```{r NA Check}
#Checking for Na's
which(is.na(orca_data_subset))
#okay so it appears we don't have any missing values. 
```

### Secondary Hypothesis - Food and Fecundity

```{r Selecting Columns 2}
orca_data_sub2 <- orca_data %>%
  select(Year, Group.ID, CK.oa, CK.tr, CM, DPERg, FLg, Ng)

#selecting 6 random groups to look at because 56 is hard to see at once
groups <- sample(1:56, 6, replace = F) #sample 6 random numbers from 1-56
groups <- as.character(groups)
for(i in 1:length(groups)){ #for each number
  if(nchar(groups[i])==1){ #if the number of characters is 1 (1, 2, 3...)
    groups[i] <- paste("G00", groups[i], sep="") #add G00 in front
  }
  else{ #if the number is not 1 (10, 11, 12...)
  groups[i] <- paste("G0", groups[i], sep="") #add G0 in front
  }
}

orca_data_sub2_6 <- orca_data_sub2 %>% #filter for the 6 groups
  filter(Group.ID == groups[1] | Group.ID == groups[2] | Group.ID == groups[3] |
         Group.ID == groups[4] | Group.ID == groups[5] | Group.ID == groups[6])

#Checking number of observations from each group
orca_data_sub2 %>% 
  group_by(Group.ID) %>% 
  tally() %>% 
  arrange(n, desc(n))

#G002, G003 have 1 observation
#G005, G0019, G022, G024, G027, G037, G046 have 2 observations
```

# We need to Remove groups with less then ?


# Plotting


## Secondary Hypothesis

```{r Exploratory Graphing 2}
#Salmon over time
ggplot(orca_data_sub2, aes(Year, CK.oa))+
  geom_point()+
  geom_line()

ggplot(orca_data_sub2, aes(Year, CK.tr))+
  geom_point()+
  geom_line()

ggplot(orca_data_sub2, aes(Year, CM))+
  geom_point()+
  geom_line()

#Lactating females over time in groups
ggplot(orca_data_sub2, aes(Year, FLg))+
  geom_point()+
  facet_wrap(~Group.ID)

ggplot(orca_data_sub2_6, aes(Year, FLg))+
  geom_point()+
  geom_line()+ #improves readability
  facet_wrap(~Group.ID)


#Number of individuals in groups
ggplot(orca_data_sub2, aes(Year, Ng))+
  geom_point()+
  facet_wrap(~Group.ID)

ggplot(orca_data_sub2_6, aes(Year, Ng))+
  geom_point()+
  geom_line()+
  facet_wrap(~Group.ID)

#pod energetic requirement
ggplot(orca_data_sub2, aes(Year, DPERg))+
  geom_point()+
  facet_wrap(~Group.ID)

ggplot(orca_data_sub2_6, aes(Year, DPERg))+
  geom_point()+
  geom_line()+
  facet_wrap(~Group.ID)
```

## Testing Normality

```{r}
ggplot(orca_data_sub2, aes(FLg))+
  geom_histogram() #histogram of FLg to assess normality
```
## Unscaling

```{r}
# value = x-mean(x) / 2sd(x)
# descale <- function(x, sd){
#   vals <- c()
#   for(i in 1:length(x)){
#     #vals[i] <- (x[i]-mean(x))/(2*sd)
#     vals[i] <- (x[i]+mean(x))*(2*sd)
#   }
#   return(vals)
# }
# 
# 
# unscaledFLg <- descale(orca_data_sub2$FLg, 0.73)
# 
# unscaledData <- orca_data_sub2 %>% 
#   select(FLg) %>% 
#   unscale()
# 
# unscaledFLg <- unscale(2*orca_data_sub2$FLg)
# unscaledCK.oa <- unscale(2*orca_data_sub2$CK.oa)
```


## Testing Assumptions (LMM)
1. Random Effects are Normally Distributed
2. Response (lactating Females) is normally distributed
3. Homogenity/Homoscedasticity of variance


Testing 1. Random Effects Normally distributed
```{r}
orca_data_sub2 %>% 
  select(CK.oa, Year, Group.ID) %>% 
  plot()
```

```{r}
orca_data_sub2 %>% 
  select(CK.oa) %>% 
  summary() 

hist(orca_data_sub2$CK.oa)
hist(orca_data_sub2$FLg) #Y is not normally distributed Transform to log?
```



###Making our Linear Model
Model #1 Treat Salmon Data as a fixed Variable. For the sake of just exploration I'm going to use just CK.oa for now

Random Variables; Year, Group ID, 
Fixed Variables : FLg, FRg, Salmon Data

#LMM with Randomized Slope
- we have random effects
- data is dependen on one another
- our data is not normal
- we expect the number of salmon to increase 'randomly' with number of lactating females

#CK.oa and FLg- simplest model just year as random effect
```{r}
library(lme4)
library(lmerTest)
CK_FLg_Year<-lmer(FLg~CK.oa+ (1|Year), data = orca_data_sub2, REML= FALSE)
summary (CK_FLg_Year)
```
__BIGNOTE___ (we dont actually know what salmon data units is so this means nothing rn :)


#CK.Oa, FLG ; Random effects: year and Group ID
```{r}
CK_FLg_Year_GroupID<-lmer(FLg~CK.oa + (1|Year) + (1|Group.ID), data = orca_data_sub2, REML= FALSE)
summary(CK_FLg_Year_GroupID)

plot(CK_FLg_Year_GroupID) # I dont know
```

I am going to do some transformations to see if I can make the plot less uggo
```{r}
# log transforming FLG 
orca_data_sub2$FLg <- log(orca_data_sub2$FLg)

CK_FLg_Year_GroupID_logtransform<-lmer(FLg~CK.oa + (1|Year) + (1|Group.ID), data = orca_data_sub2, REML= FALSE)
#looks better! 
plot(CK_FLg_Year_GroupID_logtransform)

#exponentially transforming FLG
orca_data_sub2$FLg <- exp(orca_data_sub2$FLg)


CK_FLg_Year_GroupID_exptransform<-lmer(FLg~CK.oa + (1|Year) + (1|Group.ID), data = orca_data_sub2, REML= FALSE)
#this also looks somewhat better than the original but I think the log one is a better transformation 
plot(CK_FLg_Year_GroupID_exptransform)
```


Kaylie's Attempt at Generating Data

Okay so based on Salmon data collection from 
https://www.npafc.org/wp-content/uploads/Public-Documents/2009/1199Rev1WGSA.pdf
pg - 51 (BC Chum Salmon)

For each year from 1980 to 2010 need to randomly select a salmon quantity.
Hypothetically, the abundance would go down towards the 2000's but for this beginning function I'm just gonna have it choose randomly.

Data set above:
Min value: 0.46 (million)
Max Value: 6.17 (million)

## Chum Generator
```{r}

years_measured<-(c(seq(1980,2010,1)))
#Numbers are in values of Millions of Fish
chum_generator <- function(){
  chum_density <- sample(x = seq(0.46,6.17,0.02), size = length(years_measured), replace = T) #choosing a random value between the range of 0.46 to 6.17, Replacing is True
  yearly_chum <- data.frame(years_measured, chum_density) #making data frame for each year and with the random density
    return(yearly_chum)
  }
  
chum_generator()

#Okay for reference, sample needs the x (the range to pick from) and then the size (which is how many variables its gonna pick); so don't need a for loop
  
```

## Chinook Generator
Based on https://www.npafc.org/wp-content/uploads/Public-Documents/2009/1199Rev1WGSA.pdf
pg 123
From 1980- 2008 (in millions of fish)
Min value: 0.13
Max Value: 1.27

```{r}

#Numbers are in values of Millions of Fish
chinook_generator <- function(){
  chinook_density <- sample(x = seq(0.13,1.27,0.02), size = length(years_measured), replace = T) #choosing a random value between the range of 0.46 to 6.17, Replacing is True
  yearly_chinook <- data.frame(years_measured, chinook_density) #making data frame for each year and with the random density
    return(yearly_chinook)
  }
  
chinook_generator()

```

#Generator for Salmon in DF format
```{r}
salmon<- data.frame(chinook_generator(), chum_generator())

salmon_data<-salmon %>% 
  select(years_measured,chinook_density,chum_density) %>% 
  mutate(total_salmon = chinook_density + chum_density) 
```

```{r}
hist(salmon_data$total_salmon)
```


#Pod Size Generator 
Based on 

```{r}
group_number<-seq(1,16,1)
pod <- function(){
  n<- rpois(1, lambda = 300)
  pod_size<-rmultinom(1, size = n, prob = rep(1,16)) 
  #multinorm to make sure orcas add to 300, repeat 16 times
  pod_size_by_group <- data.frame(group_number,pod_size ) #puttin gin dataframe
    return(pod_size_by_group)
  }
 
pod()

```

## Amount of Reproductive females
Based on (INSERt link)
```{r}
pod_data<- pod() %>% 
  mutate(number_repro_females = round(pod_size * 0.22,0))
#rounding upwards to get whole number

```

# Number of Lactating Females
```{r}
Lactating_females <- c()
for(i in 1:nrow(pod_data)){
  Lactating_females[i] <- rbinom(1,pod_data$number_repro_females[i], prob = 0.2)
}
Lactating_females

 pod_data_2<-pod_data %>% 
  mutate(Lactating_females = Lactating_females)

pod_data_2

```

```{r}
lactating_Orca_Data_Gen <- function(n){
  pod_data_Year <- data.frame()
  group_number<-seq(1,16,1)
  for(i in 1:n){
    pod <- function(){
      #poisson distribution to choose total population each year
      n <- rpois(1, lambda = 300)
      #multinomial distribution keeps orcas at total pop size, make 16 pods
      pod_size <- rmultinom(1, size = n, prob = rep(1,16))
      year <- i+1979
      pod_size_by_group <- data.frame(year, group_number, pod_size)
      return(pod_size_by_group)
    }

    pod_data<- pod() %>% 
      mutate(number_repro_females = round(pod_size * 0.22,0))
    Lactating_females <- c()
    for(i in 1:nrow(pod_data)){
      Lactating_females[i] <- rbinom(1,pod_data$number_repro_females[i], prob = 0.2)
    }
    pod_data_2<-pod_data %>% 
    mutate(Lactating_females = Lactating_females)
    pod_data_Year <- rbind(pod_data_Year, pod_data_2)
  }
  return(pod_data_Year)
}
year_pod_data <- lactating_Orca_Data_Gen(31)

# Did not work:
# vals <- cbind(year_pod_data, salmon_data)

# rep_salmon_data <- data.frame()
# for(i in 1:nrow(salmon_data)){
#   for(x in i:(16*i)){
#     rep_salmon_data[x,] <- salmon_data[i,]
#     print(x)
#   }
#   print(i)
# }
# 
# rep_salmon_data[1,] <- salmon_data[1,]
# rep_salmon_data[17:32,] <- salmon_data[2,]
```

MISC>
i think if salmon is a random variable it should be (1+ck.oa|Year) (because ck is dependent on y ear)

# Power Analysis

```{r}
# ------------- From Lecture -------------
power_analysis_function <- function(sample_size){
  sims <- 1000
  pvalues <- c()
  for (i in 1:sims){
  data <- data_generator(sample_size)
  model <- glm(disease_status~scale(age), family = binomial, data = data)
  pvalues[i] <- summary(model)$coefficients[2,4]
  }

  power_estimate <- length(which(pvalues < 0.01))/length(pvalues)

  return(power_estimate)
}
# 
# sample_sizes <- seq(100,200,10); power_estimates <- c()
# 
# for (i in 1:length(sample_sizes)){
#   power_estimates[i] <- power_analysis_function(sample_size = sample_sizes[i])
# }
# 
# knitr::kable(cbind(n = sample_sizes, power = power_estimates))



 #pod_size <- sample(x = seq(3,38,1), size = 16, replace = T) #Choosing from 16 total pods
```




