---
title: "Final Orca Data Project (including old hypothesis)"
output: html_notebook
---

**WARNING** Before you run code for the entire document, there is a chunk that takes over an hour to run. Please change the value of `sims` if needed to reduce run time.

# Set Up

```{r Packages}
library(here)
library(tidyverse)
library(dplyr)
library(lme4)
library(stats)
```

```{r Data Loading}
# Reading in the Data
library(here)
here()
orca_data<- read.csv(here("stredulinsky2021.csv"))

```

```{r Data Exploration}
# Exploring
head(orca_data)
glimpse(orca_data)
# 25 columns

which(is.na(orca_data))
# no NA's present
```

# Data Selection

## Column selection

- Year
- Group ID
- CK.oa (Chinook salmon ocean abundance index)
- CK.tr (Chinook salmon terminal run reconstruction abundance index) (accounting for industrial fishers affecting abundance)
- CM (chum salmon terminal run abundance index)
- DPERg (Total estimated daily prey energetic requirement of the pod)
- FLg (number of lactating females in the pod (proportion to pod size))
- Ng (number of individuals in pod)

### Subsetting the Data

```{r Selecting Columns 2}
orca_data_sub2 <- orca_data %>% #select relevant columns
  select(Year, Group.ID, CK.oa, CK.tr, CM, DPERg, FLg, Ng)
```

```{r Pod Subsets}
# Here we're selecting six random groups to look at in the graphs since
# when we facet wrap by all groups, the graphs end up small. This way
# we just graph six to see some of the data more clearly

groups <- sample(1:56, 6, replace = F) # Sample 6 random numbers from 1-56

# Modifying the selected groups to match the format in the data

# Make the groups characters
groups <- as.character(groups)
# For each group
for(i in 1:length(groups)){
  # If the number of characters is 1 (1, 2, 3...)
  if(nchar(groups[i])==1){
    groups[i] <- paste("G00", groups[i], sep="") # Add G00 in front
  }
  # If the number of characters is not 1 (10, 11, 12...)
  else{
  groups[i] <- paste("G0", groups[i], sep="") # Add G0 in front
  }
}

# Filter for the 6 groups
orca_data_sub2_6 <- orca_data_sub2 %>%
  filter(Group.ID == groups[1] | Group.ID == groups[2] | Group.ID == groups[3] |
         Group.ID == groups[4] | Group.ID == groups[5] | Group.ID == groups[6])
```

#### Low Counts for Some Groups

```{r Pod Observation Numbers}
# Checking number of observations from each group
orca_data_sub2 %>% 
  group_by(Group.ID) %>% 
  tally() %>% 
  arrange(n, desc(n))

#G002, G003 have 1 observation
#G005, G0019, G022, G024, G027, G037, G046 have 2 observations
```

Some groups have very few observations in the thirty years, and we considered removing groups with low observation counts. We ultimately didn't use this hypothesis, so we never decided on a cutoff or if we would do this.

# Plotting

```{r Exploratory Graphing 2}
# Looking at how salmon changes over time
ggplot(orca_data_sub2, aes(Year, CK.oa))+
  geom_point()+
  geom_line()

ggplot(orca_data_sub2, aes(Year, CK.tr))+
  geom_point()+
  geom_line()

ggplot(orca_data_sub2, aes(Year, CM))+
  geom_point()+
  geom_line()

# Lactating females over time in pods
ggplot(orca_data_sub2, aes(Year, FLg))+
  geom_point()+
  facet_wrap(~Group.ID)

ggplot(orca_data_sub2_6, aes(Year, FLg))+
  geom_point()+
  geom_line()+
  facet_wrap(~Group.ID)


# Number of individuals in pods
ggplot(orca_data_sub2, aes(Year, Ng))+
  geom_point()+
  facet_wrap(~Group.ID)

ggplot(orca_data_sub2_6, aes(Year, Ng))+
  geom_point()+
  geom_line()+
  facet_wrap(~Group.ID)

# Pod energetic requirement
ggplot(orca_data_sub2, aes(Year, DPERg))+
  geom_point()+
  facet_wrap(~Group.ID)

ggplot(orca_data_sub2_6, aes(Year, DPERg))+
  geom_point()+
  geom_line()+
  facet_wrap(~Group.ID)
```

## Testing Normality

```{r}
# Histogram of FLg to assess normality
ggplot(orca_data_sub2, aes(FLg))+
  geom_histogram()
```

## Unscaling

```{r}
#The data is transformed via: x = x-mean(x) / (2*sd(x)) 
descale <- function(x, sd){
  vals <- c()
  for(i in 1:length(x)){
    # Going backwards from the process the authors used
    vals[i] <- (x[i]+mean(x))*(2*sd)
  }
  return(vals)
}

# Attempts to use the unscale function (unsuccessful)
unscaledData <- orca_data_sub2 %>%
  select(FLg) %>%
  unscale()

unscaledFLg <- unscale(2*orca_data_sub2$FLg)
unscaledCK.oa <- unscale(2*orca_data_sub2$CK.oa)
```

This unscaling was unsuccessful, ultimately leading us to abandon this hypothesis and switch to generating data. We still attempted to make a linear mixed model to see what we got back. It is commented out since it

# Testing Assumptions (LMM)

1. Random Effects are Normally Distributed
2. Response (lactating Females) is normally distributed
3. Homogenity/Homoscedasticity of variance


Testing 1. Random Effects Normally distributed
```{r}
orca_data_sub2 %>% 
  select(CK.oa, Year, Group.ID) %>% 
  plot()
```

```{r}
orca_data_sub2 %>% 
  select(CK.oa) %>% 
  summary() 

hist(orca_data_sub2$CK.oa)
hist(orca_data_sub2$FLg) #Y is not normally distributed Transform to log?
```

# Making our Linear Model

Model #1 Treat Salmon Data as a fixed Variable. For the sake of just exploration we're going to use just CK.oa for now

Random Variables; Year, Group ID, 
Fixed Variables : FLg, Salmon Data

## LMM with Randomized Slope

- we have random effects
- data is dependent on one another
- our data is not normal
- we expect the number of salmon to increase 'randomly' with number of lactating females

# CK.oa and FLg- simplest model just year as random effect

```{r}
library(lme4)
library(lmerTest)
CK_FLg_Year<-lmer(FLg~CK.oa+ (1|Year), data = orca_data_sub2, REML= FALSE)
summary (CK_FLg_Year)
```

__BIGNOTE__ We don't actually know what units the salmon data is in so this is hard to interpret


#CK.Oa, FLG ; Random effects: year and Group ID

```{r}
CK_FLg_Year_GroupID<-lmer(FLg~CK.oa + (1|Year) + (1|Group.ID), 
                          data = orca_data_sub2, REML= FALSE)
summary(CK_FLg_Year_GroupID)

plot(CK_FLg_Year_GroupID)
```

We attempted to transform this plot just to see if anything made more sense

```{r}
# Log transforming FLG 
orca_data_sub2$FLg <- log(orca_data_sub2$FLg)

CK_FLg_Year_GroupID_logtransform<-lmer(FLg~CK.oa + (1|Year) + (1|Group.ID), 
                                       data = orca_data_sub2, REML= FALSE)

plot(CK_FLg_Year_GroupID_logtransform)

# Exponentially transforming FLG
orca_data_sub2$FLg <- exp(orca_data_sub2$FLg)


CK_FLg_Year_GroupID_exptransform<-lmer(FLg~CK.oa + (1|Year) + (1|Group.ID), 
                                       data = orca_data_sub2, REML= FALSE)

plot(CK_FLg_Year_GroupID_exptransform)

# Exponential transformation also looks somewhat better than the original but 
# the  the log one appears to be a better transformation
```

# Data Generation

For each year from 1980 to 2010 need to randomly select a salmon quantity.
Hypothetically, the abundance would go down towards the 2000's but for this beginning function I'm just gonna have it choose randomly.

## Creating each generator individually

### Chum Generator

This data is generated based on salmon data collection from 
https://www.npafc.org/wp-content/uploads/Public-Documents/2009/1199Rev1WGSA.pdf
pg 51 (BC Chum Salmon)
From 1980 - 2008 (in millions of fish):
Min value: 0.46 (million)
Max Value: 6.17 (million)

```{r Chum Generator}
years_measured<-(c(seq(1980,2010,1)))
# Values are in of millions of fish
chum_generator <- function(){
  # Choosing a random value between the range of 0.46 to 6.17, 
  # replacement is true
  chum_abundance <- sample(x = seq(0.46,6.17,0.02), 
                         size = length(years_measured), 
                         replace = T) 
  # Making data frame for each year and with the random density
  yearly_chum <- data.frame(years_measured, chum_abundance)
    return(yearly_chum)
  }
```

### Chinook Generator

Based on https://www.npafc.org/wp-content/uploads/Public-Documents/2009/1199Rev1WGSA.pdf
pg 123
From 1980 - 2008 (in millions of fish)
Min value: 0.13
Max Value: 1.27

```{r Chinook Generator}
# Values are in millions of Fish
chinook_generator <- function(){
  # Choosing a random value between the range of 0.13 to 1.27, 
  # replacement is true
  chinook_abundance <- sample(x = seq(0.13,1.27,0.02), 
                            size = length(years_measured), 
                            replace = T)
  # Making data frame for each year and with the random density
  yearly_chinook <- data.frame(years_measured, chinook_abundance) 
    return(yearly_chinook)
  }
```

### Generator for Salmon in DF format
We needed to change the format so we could combine our data into one data frame and graph it as a histogram.
```{r Salmon Generation Dataframe}
salmon <- data.frame(chinook_generator(), chum_generator())

# Combining Chinook and Chum Generator
salmon_data <- salmon %>% 
  select(years_measured, chinook_abundance, chum_abundance) %>% 
  mutate(total_salmon = chinook_abundance + chum_abundance) 
```

```{r Salmon Histogram}
# Assessing normality of salmon
hist(salmon_data$total_salmon)
```


### Pod Size Generator 

Based on information from https://media.fisheries.noaa.gov/dam-migration/2019_sars_alaska_killer_whale_-_eastern_north_pacific_northern_resident.pdf
Approximate Total Population: ~300
Typical Pod Count: ~16

```{r Pod Size Generator}
# Create groups 1-16
group_number<-seq(1,16,1)

pod <- function(){
  # Allow total number of orcas to vary each year
  n <- rpois(1, lambda = 300)
  pod_size < -rmultinom(1, size = n, prob = rep(1,16)) 
  # Multinomial to create 16 pods of total n orcas
  pod_size_by_group <- data.frame(group_number, pod_size) 
  # Putting the values in dataframe
  return(pod_size_by_group)
}
```

### Amount of Reproductive females

Based on www.sararegistry.gc.ca/status/status_e.cfm

```{r Reproductive Females Calculation}
pod_data <- pod() %>% 
  # Multiply pod size by the average percent of reproductive age females in a 
  # pod and round it to a whole number
  mutate(number_repro_females = round(pod_size * 0.22,0))

```

### Number of Lactating Females

This ultimately was not used in the final generator. The chance a reproductive age female orca had a child in a given year was reported as 20% in a CBC article. However this ignored any effects of salmon, which we needed to include.

```{r Lactating Females Calculation}
# Create a data frame to hold lactating female count
Lactating_females <- c()

# Loop for the total number of pods over all years
for(i in 1:nrow(pod_data)){
  # Use a binomial to select reproductive females to be lactating
  Lactating_females[i] <- rbinom(1,pod_data$number_repro_females[i], prob = 0.2)
}

# Add lactating females to the data frame
pod_data_2 <- pod_data %>% 
  mutate(Lactating_females = Lactating_females)
```

## Final Data Generator

```{r Data Generator}
data_Gen <- function(beta1, beta2){
  pod_data_Year <- data.frame()
  
  # ----- Salmon Data -----
  # Generate salmon data in a dataframe
  salmon <- data.frame(chinook_generator(), chum_generator())
  
  # Format the data frame and add total salmon
  salmon_data <- salmon %>% 
    select(years_measured, chinook_abundance, chum_abundance) %>% 
    mutate(total_salmon = chinook_abundance + chum_abundance) 
  
  # Create a data frame to math the format of salmon to orca data
  rep_salmon_data <- data.frame(chinook_abundance = c(), 
                                chum_abundance = c(), 
                                total_salmon = c())
  
  # Setting the counters to one
  x <- 1
  i <- 1
  
  # Go through the number rows of the orca data
  while(x < 496){
    #for each 16 rows of an empty data frame, add the corresponding year 
    #from salmon data
    rep_salmon_data[c(x:(x+16)),1] <- salmon_data[i,2]
    rep_salmon_data[c(x:(x+16)),2] <- salmon_data[i,3]
    rep_salmon_data[c(x:(x+16)),3] <- salmon_data[i,4]
    x <- x+16 #increase rows in repeated salmon data
    i <- i+1 #increase row in salmon data
  }
  
  #remove a repeated final column
  rep_salmon_data <- rep_salmon_data[-c(497),]
  
  # Add column names
  colnames(rep_salmon_data) <- c("chinook_abundace", 
                                 "chum_abundance", 
                                 "total_salmon")
  
  # ----- Orca Data Generation -----
  
  # Each pod is given a number
  group_number<-seq(1,16,1)
  # Total number of years (1980-2010)
  n <- 31
  
  # Generating pods for each year
  for(i in 1:n){
    pod <- function(){
      # Poisson distribution to choose total population each year
      s <- rpois(1, lambda = 300)
      # Multinomial distribution keeps orcas at total pop size, make 16 pods
      pod_size <- rmultinom(1, size = s, prob = rep(1,16))
      year <- i+1979
      pod_size_by_group <- data.frame(year, group_number, pod_size)
      return(pod_size_by_group)
    }

    # Generating lactating females
    pod_data <- pod() # Call pod function above to make pods
    Lactating_females <- c() # Create a vector to hold lactating females
    
    # Calculate lactating females
    for(i in 1:nrow(pod_data)){
      # Betas - makes lactating females dependent on salmon + pod size
      Lactating_females[i] <- rpois(1,
                                lambda = beta1*(pod_data$pod_size[i])+ 
                                         beta2*rep_salmon_data$total_salmon[i])
      # This poisson determines lactating females from pod size + additional 
      # lactating females with w/ more salmon
    }
    
    # Add lactating females column
    pod_data_2 <- pod_data %>% 
      mutate(lactating_females = Lactating_females)
    
    # Bind data frames to an overall data frame
    pod_data_Year <- rbind(pod_data_Year, pod_data_2)
  }

  # Combining the dataframe
  gen_Data <- pod_data_Year %>%
    # Add the salmon columns
    mutate(chinook_abundance = c(rep_salmon_data[,1])) %>%
    mutate(chum_abundance = c(rep_salmon_data[,2])) %>%
    mutate(total_salmon = c(rep_salmon_data[,3]))
  
  return(gen_Data)
}
```

# Power Analysis

**WARNING** The below code takes over an hour (roughly 3-4hrs) to run. Change simulations to 100 for it to run in a few minutes if you don't have the time to wait.

```{r Power Analysis, warning = F}
# Set parameters
params <- expand.grid(beta1 = seq(0.2,0.24,0.01), 
                      beta2 = c(0.0000001,0.00001,0.0001,0.001,0.01,0.1))

# Beta needs to be a vector with two entries - 1st: beta1, 2nd: beta2
power_poisson <- function(beta){
  # Number of models to create for each set of parameters
  sims <- 1000
  
  # Create vectors/variables to hold values
  pValSalmon <- c()
  pValPod <- c()
  SalmonEffect <- c()
  PodEffect <- c()
  models <- NULL
  predictions <- NULL
  Intercept <- c()
  
  # Loop for the number of simulations
  for(i in 1:sims){
    # Generate the orca data for 31 years (1980-2010)
    data <- data_Gen(beta[1], beta[2])
    # Make a glmm with lactating females as (y), and fixed effects total_salmon 
    # and pod_size as (x). Random effects are year and group_number
    model <- glmer(lactating_females~total_salmon+pod_size+
                   (1|group_number)+(1|year), 
                 family = poisson, data)
    models[[i]] <- model # Saving the models to return if needed
    
    # Predict the number of lactating females from the model, and save the
    # total salmon, year, pod_size, group_number, and model index
    predictions[[i]] <- cbind(lact_fem_predict=
                                predict(models[[i]], 
                                        newdata = data.frame(
                                          total_salmon=data$total_salmon,
                                          pod_size = data$pod_size,
                                          year=data$year,
                                          group_number=data$group_number)), 
                              total_salmon = data$total_salmon,
                              year = data$year,
                              pod_size = data$pod_size,
                              group_number = data$group_number,
                              index = i)
    
    # Saving the effect of salmon and pod size
    SalmonEffect[i] <- summary(model)$coefficient[2,1]
    PodEffect[i] <- summary(model)$coefficient[3,1]
    
    # Saving the intercept
    Intercept[i] <- summary(model)$coefficient[1,1]
    
    # Saving model significance for each beta
    pValSalmon[i] <- summary(model)$coefficient[2,4]
    pValPod[i] <- summary(model)$coefficient[3,4]
  }
  
  # Calculating how many models were significant out of the generated models
  powEstSalmon <- length(which(pValSalmon<0.05))/length(pValSalmon)
  powEstPod <- length(which(pValPod<0.05))/length(pValPod)
  
  return(list(SalmonEffect, PodEffect, Intercept,
              powEstSalmon, powEstPod, do.call(rbind, predictions)))
}

# Create vectors to save values
PowersSalmon <- c()
powEstPod <- c()

# For each row in the parameters, run the power analysis and save 
# the power estimates
for (i in 1:nrow(params)){
  beta <- c(params[i,1], params[i,2])
  result <- power_poisson(beta)
  PowersSalmon[i] <- result[[4]]
  powEstPod[i] <- result[[5]]
}

# Creating a dataframe of powers at the various beta levels
cbind(PowersSalmon, powEstPod, params)

seq(min(as.data.frame(result[[6]])$total_salmon),
    max(as.data.frame(result[[6]])$total_salmon), length = 1000) -> salmons

seq(min(as.data.frame(result[[6]])$pod_size),
    max(as.data.frame(result[[6]])$pod_size), length = 1000) -> pods

```

### Graphing the model slopes

```{r Model Plotting}
# Create variables to save the lines
salmon_lines <- NULL
pod_lines <- NULL

# Create lines where one fixed effect is constant using the slopes and
# intercepts from all of the models
for (i in 1:length(result[[3]])){
  # Create a data frame of all salmon model intercepts and slopes while 
  # keeping pod size constant
  lines1 <- as.data.frame(cbind(salmons = salmons)) %>%
    mutate(lact_prediction = result[[3]][i]+
             result[[1]][i]*salmons)
  lines1 <- cbind(lines1, index_model = i)
  salmon_lines <- rbind(salmon_lines, lines1)
  
  # Create a data frame of all pod size model intercepts and slopes while 
  # keeping salmon abundance constant
  lines2 <- as.data.frame(cbind(pods = pods)) %>%
    mutate(lact_prediction = result[[3]][i]+
             result[[2]][i]*pods)
  lines2 <- cbind(lines2, index_model = i)
  pod_lines <- rbind(pod_lines, lines2)
}

# Generate plots with the models of salmon and pod effects on the predicted 
# lactating females
as.data.frame(salmon_lines) %>%
  ggplot(aes(x = salmons, y = lact_prediction, group = index_model))+
  geom_line(color = "gray")+
  geom_line(aes(x = mean(salmons), y = mean(lact_prediction)), colour = "blue")+
  labs(x = "Salmon Abundance (millions of fish)", 
       y ="Predicted Number of Lactating Females per Pod" , 
       title = "Effect of Salmon Abundance on Lactating Females")+
  theme_bw()

as.data.frame(pod_lines) %>% 
  ggplot(aes(x = pods, y = lact_prediction, group = index_model))+
  geom_line(color = "gray")+
  labs(x = "Size of Pod", 
       y = "Predicted Number of Lactating Females", 
       title = "Pod Size on Number of Lactating Females")+
  theme_bw()
```




